Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-07-04 survey/survey.tex
Rewrote parts of the intro. Decided not to move the text, as most of it could be kept as is without giving away the complete punchline. Though some of it might still bear repeating in the conclusion.

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016}
introduced the~$\Pi$ family of reversible languages based on type
isomorphisms and commutative semiring identities. As motivated above,
the starting point for this development is the physical principle of
``conservation of
information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative} and
the family of languages is designed to embrace that principle by
requiring all computations to preserve information.

The fragment without
recursive types is universal for reversible boolean
circuits~\cite{James:2012:IE:2103656.2103667} and the extension with
recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete
reversible language~\cite{James:2012:IE:2103656.2103667,rc2011}. While
at first sight, $\Pi$ might appear \emph{ad hoc}, it really arises
naturally from an ``extended'' view of the Curry-Howard
correspondence~\cite{Carette2016}: rather than looking at mere
\emph{inhabitation} as the main source of analogy between logic and
computation, \emph{type equivalences} becomes the source of analogy.
This allows one to see an analogy between algebra and reversible
computation.  Furthermore, this works at multiple levels: that of
$1$-algebra (types form a semiring under isomorphism), but also
$2$-algebra (types and equivalences form a weak Rig Groupoid).  In
other words, by taking ``weak Rig Groupoid'' as the starting
semantics, one naturally gets $\Pi$ as the syntax for the language of
proofs of isomorphisms -- in the same way that many terms of the
$\lambda$-calculus arise from Cartesian Closed Categories.

% \amr{We should also have a quick survey of other works in reversible
% computation? Most of the other work tries to start from well known
% models of computation or well known programming languages, and then
% adapts them to be reversible. Pi is different in that it starts from
% the semantic implications of reversibility. Then, because it adds
% types, rather than control-flow (or even data-flow) as its next layer
% of ``understanding,'' this leads to equivalences, isomorphisms, etc.
% This then leads, quite naturally, to finding that the ``proof
% language'' of semirings (and Rig Groupoids at level 2) is actually a
% programming language. And it is Pi. This is a neat twist on
% Curry-Howard because CH is about \textbf{inhabitation} only. But with
% ``conservation of information'' as the basis, a different kind of
% correspondence arises; in fact, this one may well be an actual
% isomorphism.}

% \jc{I think it would be fair to have a \textit{quick} survey
% of other works. The most important aspect would be to draw out that
% they all start from known models of computation and then attempt to
% make them reversible.  Our originality is to start from the semantics
% of reversibility (even further back, preservation of information)
% and derive a programming language purposefully designed to denote
% exactly that. Having said that, I am not in a good position to
% write such a survey, being the least knowledgeable of all 3 of us
% of this general area.}

>>>>

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016}
introduced the~$\Pi$ family of typed reversible languages.  As motivated above,
the starting point for this development is the physical principle of
``conservation of
information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative} and
the family of languages is designed to embrace that principle by
requiring all computations to preserve information.

The fragment without
recursive types is universal for reversible boolean
circuits~\cite{James:2012:IE:2103656.2103667} and the extension with
recursive types and trace
operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete
reversible language~\cite{James:2012:IE:2103656.2103667,rc2011}. While
at first sight, $\Pi$ too might appear \emph{ad hoc}, it really arises
naturally from an ``extended'' view of the Curry-Howard
correspondence~\cite{Carette2016}: rather than looking at mere
\emph{inhabitation} as the main source of analogy between logic and
computation, \emph{type equivalences} becomes the source of analogy.
Taking inspiration from the fact that many terms of the $\lambda$-calculus
arise from Cartesian Closed Categories including, most importantly,
a variety of propositional equalities and computation rules,
allows us to pursue that analogy further. Some of the details
of this development will be motivated and explained in the
present paper.
